---
title: "Data Analysis"
author: "Mia Isaacs"
date: "2025-03-27"
output: github_document
---

### load packages

```{r}
library(tidyverse)
library(Biostrings)
library(Pviz)
library(Gviz)
library(stringr)
library(GenomicRanges)
library(preprocessCore) 
library(mclust) 
library(lme4)
library(pROC)
library(DescTools)
library(forcats)
library(scales)
library(viridis)
library(ggeffects)
```

### view dataset

```{r}
dengueIgG_all = read_csv("data/Dengue_data_analysis_IgG.csv") |> 
  janitor::clean_names()

view(head(dengueIgG_all, 10))
```

### remove unncessary variables
```{r}
dengueIgG_clean <- dengueIgG_all |> 
  select(-accession, -p_val_acute_conv, -p_adjust_acute_conv, -p_val_acute_nc, -p_adjust_acute_nc, -p_val_conv_nc,
         -p_adjust_conv_nc, -x10, -acute, -conv, -nc, -row_total_x, -control, -d1_acute, -d1_conv, -d2_acute,
         -d2_conv, -d3_acute, -d3_conv, -d3d4_acute, -d3d4_conv, -d4_acute, -d4_conv,z_conv, -zd_conv, -zdyfv_conv,
         -zyfv_conv, -row_total_y, -z_conv)

head(dengueIgG_clean)
```

### check for missing values (all good!)
```{r}
sum(is.na(dengueIgG_clean))
```

### convert to long format for easier manipulation
```{r}
dengueIgG_long <- dengueIgG_clean |> 
  pivot_longer(cols = matches("uwa_"),
               names_to = "sample",
               values_to = "fluorescence")

head(dengueIgG_long)
```

### add log transformed fluorescence values for future analysis
```{r}
dengueIgG_transform <- dengueIgG_long |> 
  mutate(fluorescence_log = log(fluorescence))

head(dengueIgG_transform)
```

## prepare data for quantile normalization

### extract serotype from sample name
```{r}
dengueIgG_serotype <- dengueIgG_transform |> 
  mutate(serotype = str_extract(sample, "_(d[1-4]d[1-4]|d[1-4]|z|zd|zdyfv|zyfv|nc)_") |>  
           str_replace_all("_", "")) |> 
  arrange(sample)
```

### visualize distributions of fluorescence values across serotypes
```{r}
ggplot(dengueIgG_serotype, aes(x = serotype, y = fluorescence)) +
  geom_boxplot() +
  ggtitle("Original Fluorescence Distributions by Serotype")
```

fluorescence values appear to be similarly distributed across all serotypes, except for the negative control. despite the similarities, i will complete quantile normalization to be sure that my data is properly aligned.

### normalize across all serotypes
```{r}
# extract fluorescence values as a matrix
fluorescence_matrix <- as.matrix(dengueIgG_serotype |> select(fluorescence))

# perform quantile normalization across ALL samples together
fluorescence_norm <- normalize.quantiles(fluorescence_matrix)

# add normalized values back to the full dataset
dengueIgG_serotype$fluorescence_norm <- as.vector(fluorescence_norm)

head(dengueIgG_serotype)
```

### summarize fluorescence values to normalized values to observe any differences
```{r}
original_summary <- dengueIgG_serotype |> 
  summarize(
    min = min(fluorescence, na.rm = TRUE),
    q1 = quantile(fluorescence, 0.25, na.rm = TRUE),
    median = median(fluorescence, na.rm = TRUE),
    mean = mean(fluorescence, na.rm = TRUE),
    q3 = quantile(fluorescence, 0.75, na.rm = TRUE),
    max = max(fluorescence, na.rm = TRUE),
    sd = sd(fluorescence, na.rm = TRUE)
  )

normalized_summary <- dengueIgG_serotype |> 
  summarize(
    min = min(fluorescence_norm, na.rm = TRUE),
    q1 = quantile(fluorescence_norm, 0.25, na.rm = TRUE),
    median = median(fluorescence_norm, na.rm = TRUE),
    mean = mean(fluorescence_norm, na.rm = TRUE),
    q3 = quantile(fluorescence_norm, 0.75, na.rm = TRUE),
    max = max(fluorescence_norm, na.rm = TRUE),
    sd = sd(fluorescence_norm, na.rm = TRUE)
  )

print(original_summary)
print(normalized_summary)
```

### visualize fluorescence vs. normalized values in density plots
```{r}
ggplot(sample_n(dengueIgG_serotype, 5000), aes(x = fluorescence)) +
  geom_density() +
  ggtitle("Fluorescence Density (Original)")

ggplot(sample_n(dengueIgG_serotype, 5000), aes(x = fluorescence_norm)) +
  geom_density() +
  ggtitle("Fluorescence Density (Normalized)")
```

according to the summary statistics and density plots, the normalized fluorescence values are identical to the original fluorescence values. therefore, quantile normalization was not necessary for this dataset but will be maintained until completion of the analysis. (may not be included in report)

### extract necessary details from annotations
```{r}
dengueIgG_serotype <- dengueIgG_serotype |> 
  mutate(
    sample_id = str_extract(sample, "(?<=_)[0-9]+(_[0-9]+)?"), 
    patient_id = str_extract(sample_id, "^[0-9]+"),
    timepoint = if_else(str_detect(sample_id, "_"), str_extract(sample_id, "(?<=_)[0-9]+"), NA_character_)
  )

head(dengueIgG_serotype)
tail(dengueIgG_serotype)
```

## separation into serotype-specific and acute vs. conv data frames
```{r, d1 acute}
normal_d1_acute <- dengueIgG_serotype |> 
  filter(serotype == "d1", str_detect(sample, "acute"))
```

```{r, d1 conv}
normal_d1_conv <- dengueIgG_serotype |> 
  filter(serotype == "d1", str_detect(sample, "conv"))
```

```{r, d2 acute}
normal_d2_acute <- dengueIgG_serotype |> 
  filter(serotype == "d2", str_detect(sample, "acute"))
```

```{r, d2 conv}
normal_d2_conv <- dengueIgG_serotype |> 
  filter(serotype == "d2", str_detect(sample, "conv"))
```

```{r, d3 acute}
normal_d3_acute <- dengueIgG_serotype |> 
  filter(serotype == "d3", str_detect(sample, "acute"))
```

```{r, d3 conv}
normal_d3_conv <- dengueIgG_serotype |> 
  filter(serotype == "d3", , str_detect(sample, "conv"))
```

```{r, d4 acute}
normal_d4_acute <- dengueIgG_serotype |> 
  filter(serotype == "d4", , str_detect(sample, "acute"))
```

```{r, d4 conv}
normal_d4_conv <- dengueIgG_serotype |> 
  filter(serotype == "d4", , str_detect(sample, "conv"))
```

```{r, d3d4 acute}
normal_d3d4_acute <- dengueIgG_serotype |> 
  filter(serotype == "d3d4", str_detect(sample, "acute"))
```

```{r, d3d4 conv}
normal_d3d4_conv <- dengueIgG_serotype |> 
  filter(serotype == "d3d4", str_detect(sample, "conv"))
```

the remaining serotype groups have only convalescent samples!

```{r, z}
normal_z <- dengueIgG_serotype |> 
  filter(serotype == "z")
```

```{r, zd}
normal_zd <- dengueIgG_serotype |> 
  filter(serotype == "zd")
```

```{r, zdyfv}
normal_zdyfv <- dengueIgG_serotype |> 
  filter(serotype == "zdyfv")
```

```{r, zyfv}
normal_zyfv <- dengueIgG_serotype |> 
  filter(serotype == "zyfv")
```

```{r, nc}
normal_nc <- dengueIgG_serotype |> 
  filter(serotype == "nc")
```

## determine threshold for significance

### write function for em threshold
```{r}
em_threshold <- function(mfi_values) {
  # fit a Gaussian mixture model with 2 components (noise and signal)
  fit <- Mclust(mfi_values, G = 2)
  
  # get parameters of the two components
  means <- fit$parameters$mean
  sds <- sqrt(fit$parameters$variance$sigmasq)
  
  # identify noise distribution (lower mean component)
  noise_component <- which.min(means)
  mean_noise <- means[noise_component]
  sd_noise <- sds[noise_component]
  
  # calculate threshold (mean + 2SD of noise distribution)
  threshold <- mean_noise + 2 * sd_noise
  return(threshold)
}
```

### apply function to dataset
```{r}
processed_data <- dengueIgG_serotype |> 
  group_by(sample_id) |>
  mutate(
    threshold = em_threshold(fluorescence_norm),
    mfi_adjusted = pmax(fluorescence_norm - threshold, 0),
    hit_yn = if_else(mfi_adjusted > 0, 1, 0)
  ) |> 
  ungroup()
```

### write function for visualization of MFI with EM threshold
```{r}
plot_threshold_distribution <- function(sample_data) {
  sample_data |> 
    ggplot(aes(x = fluorescence_norm)) +
    geom_histogram(aes(y = ..density..), bins = 50, fill = "lightblue") +
    geom_density(color = "darkblue", linewidth = 1) +
    geom_vline(aes(xintercept = threshold), color = "red", linetype = "dashed", linewidth = 1) +
    labs(title = "MFI Distribution with E-M Threshold",
         x = "MFI Value",
         y = "Density") +
    theme_minimal()
}
```

### apply to sample as an example
```{r}
sample_1_data <- processed_data |>  
  filter(sample_id == unique(processed_data$sample_id)[1])
plot_threshold_distribution(sample_1_data)
```

### view processed dataset
```{r}
view(head(processed_data, 100))
```

### import data from geneious prime
```{r}
denv1_protein_regions <- read.csv("data/probe_annotations.csv") |> 
  janitor::clean_names()

probe_annotations <- denv1_protein_regions |> 
  mutate(document_name = as.character(document_name)) |>
  filter(
    !grepl("source dengue virus", name, ignore.case = TRUE),
    !grepl("Search Hit", name, ignore.case = TRUE),
    !grepl("POLY CDS", name, ignore.case = TRUE),
    !grepl("polyprotein", name, ignore.case = TRUE)
  ) |> 
  group_by(document_name) |> 
  distinct(document_name, .keep_all = TRUE) |> 
  ungroup() |> 
  mutate(
    protein = str_extract(name, "NS\\d+[A-Z]?|[CEP][^ ]*|prM"),
    sequence = str_replace_all(sequence, "\\.\\.\\.", "")
  ) |> 
  select(
    probe = document_name,
    sequence,
    start = min_original_sequence,
    end = max_original_sequence,
    protein
  )

view(head(probe_annotations, 10))
```

### merge geneious data with processed data
```{r}
final_df_long <- probe_annotations |> 
  left_join(processed_data, by = c("sequence" = "probe_sequence"), relationship = "many-to-many") |> 
  select(-annotation, -sample)

view(head(final_df_long,100))
```


## normalize data and create heatmaps

### add protein id column
```{r}
processed_data <- processed_data |> 
  mutate(protein = str_extract(annotation, "(?<=protein_id=)[A-Z]{2}_\\d+\\.\\d+"))

view(head(processed_data, 100))
```

### create data frame with total peptides per virus
```{r}
total_peptides_per_virus <- processed_data |>
  group_by(serotype, timepoint) |>
  summarise(total_peptides = n_distinct(probe_sequence, na.rm = TRUE), .groups = "drop")

print(total_peptides_per_virus)
```

### count reactive peptides per serotype & timepoint
```{r}
reactive_peptide_counts <- processed_data |>
  filter(hit_yn == 1) |> 
  distinct(serotype, timepoint, probe_sequence) |>
  group_by(serotype, timepoint) |> 
  summarize(
    reactive_peptides = n(),
    .groups = "drop"
  )

print(head(reactive_peptide_counts))
```

### normalize number of hits within each serotype & timepoint
```{r}
normalized_hit_counts <- reactive_peptide_counts |> 
  left_join(total_peptides_per_virus, by = c("serotype", "timepoint"))   |> 
  mutate(
    normalized_count = (reactive_peptides / total_peptides) * 100
  ) |> 
  filter(!is.na(timepoint))

print(head(normalized_hit_counts))
```

### order data for plotting
```{r}
normalized_hit_counts$serotype <- fct_rev(factor(normalized_hit_counts$serotype, 
                                                 levels = c("d1", "d2", "d3", "d4",
                                                            "d3d4", "z", "zd", "zyfv",
                                                            "zdyfv", "nc")))
```

### create heatmap by serotype
```{r}
ggplot(normalized_hit_counts, aes(x = timepoint, y = serotype, fill = normalized_count)) +
  geom_tile() +
  scale_fill_gradientn(
    colours = rev(heat.colors(100)), 
    labels = function(x) paste0(round(x, 1), "%")
  ) +
  labs(
    title = "Heatmap of Normalized Peptide Counts",
    x = "Timepoint",
    y = "Serotype",
    fill = "Normalized Count (%)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## make heatmap by protein

### count peptides per protein
```{r}
total_peptides_per_protein <- final_df_long |>
  group_by(protein, timepoint) |>
  summarise(
    total_peptides = n_distinct(sequence, na.rm = TRUE),
    .groups = "drop"
  )

print(total_peptides_per_protein)
```

### count reactive peptides per protein
```{r}
reactive_peptide_counts_protein <- final_df_long |>
  filter(hit_yn == 1) |> 
  distinct(protein, timepoint, sequence) |>
  group_by(protein, timepoint) |> 
  summarise(
    reactive_peptides = n(),
    .groups = "drop"
  )

print(head(reactive_peptide_counts_protein))
```

### normalize hit counts
```{r}
normalized_hit_counts_protein <- reactive_peptide_counts_protein |> 
  left_join(total_peptides_per_protein, by = c("protein", "timepoint")) |> 
  mutate(
    normalized_count = (reactive_peptides / total_peptides) * 100
  ) |> 
  filter(!is.na(timepoint))  # remove any missing timepoints just in case

print(head(normalized_hit_counts_protein))

```

### set order of proteins by genome
```{r}
genome_protein_order <- c("C", "prM", "E", "NS1", "NS2A", "NS2B", "NS3", "NS4A", "NS4B", "NS5")

normalized_hit_counts_protein$protein <- factor(
  normalized_hit_counts_protein$protein,
  levels = genome_protein_order
) |> 
  fct_rev()
```

### plot heatmap of protein by timepoint
```{r}
ggplot(normalized_hit_counts_protein, aes(x = timepoint, y = protein, fill = normalized_count)) +
  geom_tile(color = "white") +
  scale_fill_gradientn(
    colours = rev(heat.colors(100)),
    labels = function(x) paste0(round(x, 1), "%"),
    na.value = "grey90"
  ) +
  labs(
    title = "Heatmap of Normalized Peptide Counts by Protein",
    x = "Timepoint",
    y = "Protein",
    fill = "Normalized Count (%)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )
```


## making plots for reactivity across genome for each serotype and timepoint

### create table for protein positions
```{r}
protein_positions <- tibble(
  protein = c("C", "prM", "E", "NS1", "NS2A", "NS2B", "NS3", "NS4A", "NS4B", "NS5"),
  start = c(1, 115, 281, 776, 1128, 1346, 1476, 2095, 2245, 2494),
  end = c(114, 280, 775, 1127, 1345, 1475, 2094, 2244, 2493, 3392)
)
```

### filter for reactive probes only
```{r}
figure3_df <- final_df_long %>% 
  mutate(midpoint = (start + end) / 2) %>% 
  filter(hit_yn == 1) %>% 
  mutate(
    serotype_timepoint = paste(serotype, timepoint, sep = "_")
  )
```

# NEEDS MAJOR HELP, IN PROGRESS
### create plot
```{r}
ggplot(figure3_df, aes(x = midpoint, y = serotype_timepoint, color = serotype)) +
  geom_point(aes(size = fluorescence_log), alpha = 0.7) +

  # protein background regions
  geom_rect(data = protein_positions,
            aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf),
            fill = "gray90", inherit.aes = FALSE, alpha = 0.2) +

  # protein names
  geom_text(data = protein_positions,
            aes(x = (start + end) / 2, y = 0, label = protein),
            inherit.aes = FALSE, size = 3, vjust = -4) +

  scale_color_manual(values = c(
    "d1" = "red",
    "d2" = "blue",
    "d3" = "green",
    "d4" = "purple",
    "d3d4" = "orange",
    "z" = "pink",
    "zd" = "turquoise",
    "zyfv" = "gold",
    "zdyfv" = "gray"
  )) +

  theme_bw() +
  theme(
    strip.text.y = element_text(angle = 0),
    panel.spacing = unit(1, "lines"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title.y = element_blank(),
    legend.position = "bottom"
  ) +

  labs(
    x = "Genome Position (Midpoint AA)",
    y = NULL,
    color = "Serotype",
    size = "Log Fluorescence Intensity"
  )
```

```{r}
# Subset DENV1-4 and NC
figure3_denv_df <- figure3_df %>%
  filter(serotype %in% c("d1", "d2", "d3", "d4", "nc"))
```

```{r}
# Define custom color palette
serotype_colors <- c(
  "d1" = "#D7191C",
  "d2" = "#2C7BB6",
  "d3" = "#31A354",
  "d4" = "#756BB1",
  "nc" = "black"
)

# Create the plot
p1 <- ggplot(figure3_denv_df, aes(x = midpoint, y = fluorescence_log, color = serotype)) +
  
  # Protein background regions
  geom_rect(data = protein_positions,
            aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf),
            fill = "gray90", inherit.aes = FALSE, alpha = 0.4) +
  
  # Points: Fluorescence vs genome position
  geom_point(size = 0.8, alpha = 0.7) +
  
  # Protein names
  geom_text(data = protein_positions,
            aes(x = (start + end) / 2, y = max(figure3_denv_df$fluorescence_log, na.rm = TRUE) + 0.3, label = protein),
            inherit.aes = FALSE, size = 3, vjust = 0) +
  
  scale_color_manual(values = serotype_colors) +
  
  theme_bw(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(size = 10),
    panel.spacing = unit(1, "lines"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.y = element_text(size = 12),
    axis.title.x = element_text(size = 12),
    legend.position = "bottom",
    legend.box = "vertical",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8)
  ) +
  
  labs(
    x = "Genome Position (Midpoint AA)",
    y = "Log Fluorescence Intensity",
    color = "Serotype",
    title = "DENV1-4 and NC"
  ) +
  
  # Facet by serotype and timepoint
  facet_grid(serotype ~ timepoint, scales = "free_x", space = "free_x")

p1
```

```{r}
# Subset other serotypes
figure3_other_df <- figure3_df %>%
  filter(!serotype %in% c("d1", "d2", "d3", "d4", "nc"))
```

# RETURN TO LATER, NEED BREAK





## linear mixed effects model

### fit basic model
```{r}
# make sure timepoint is ordered/numeric
processed_data$timepoint <- as.numeric(as.character(processed_data$timepoint))

# fit mixed effects model
basic_model <- lmer(
  fluorescence_norm ~ timepoint + (1 | patient_id), 
  data = processed_data
)

summary(basic_model)
```

### fit basic model in log scale (?)
```{r}
basic_log_model <- lmer(
  fluorescence_log ~ timepoint + (1 | patient_id), 
  data = processed_data
)

summary(basic_log_model)
```

### fit model with interaction by serotype
```{r}
serotype_model <- lmer(
  fluorescence_norm ~ timepoint * serotype + (1 | patient_id), 
  data = processed_data
)

summary(serotype_model)
```

### fit model with interaction in log scale 
```{r}
serotype_model_log <- lmer(
  fluorescence_log ~ timepoint * serotype + (1 | patient_id), 
  data = processed_data
)

summary(serotype_model_log)
```

### plot responses by serotype across timepoints
```{r}
# get predicted log fluorescence values over time for each serotype
preds <- ggpredict(serotype_model_log, terms = c("timepoint", "serotype"))

ggplot(preds, aes(x = x, y = predicted, color = group, fill = group)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, color = NA) +
  labs(
    title = "Predicted Antibody Dynamics Over Time by Serotype",
    x = "Timepoint",
    y = "Log Fluorescence",
    color = "Serotype",
    fill = "Serotype"
  ) +
  theme_minimal(base_size = 14)
```



### calculate intensity and breadth
```{r}
# summarize intensity and breadth for each virus in each sample
varscore_components <- dengueIgG_serotype |> 
  group_by(sample, serotype) |> 
  summarize(
    intensity = sum(adjusted_reactivity, na.rm = TRUE), # sum of adjusted reactivities
    breadth = mean(adjusted_reactivity > 0, na.rm = TRUE), # proportion of positive peptides
    .groups = "drop"
  )
```

### normalize scores
```{r}
# normalize intensity and breadth by maximum values across all samples
varscore_components <- varscore_components |> 
  group_by(sample) |> 
  mutate(
    norm_intensity = ifelse(max(intensity, na.rm = TRUE) == 0, 0, intensity / max(intensity, na.rm = TRUE)),
    norm_breadth = ifelse(max(breadth, na.rm = TRUE) == 0, 0, breadth / max(breadth, na.rm = TRUE))
  ) |> 
  ungroup() |> 
  mutate(VARscore = norm_intensity * 0.7 + norm_breadth * 0.3)

print(head(varscore_components))
print(tail(varscore_components))
```

### validate VARscores
```{r}
virus_pattern <- "_(d[1-4]d[1-4]|d[1-4]|z|zd|zdyfv|zyfv|nc)_"

varscore_components <- varscore_components |> 
  mutate(
    is_positive = ifelse(str_detect(serotype, virus_pattern), 1, 0)
  )

print(head(varscore_components))

# Step 8: ROC Analysis
roc_curve <- roc(varscore_components$is_positive, varscore_components$VARscore)

# Print and plot ROC curve
print(roc_curve)
plot(roc_curve, main = "ROC Curve for VARscore")
```




### preparing data for creation of VARscore
```{r}
dengueIgG_serotype <- dengueIgG_serotype |> 
  mutate(
    sample_id = str_extract(sample, "(?<=_)[0-9]+(_[0-9]+)?"), 
    patient_id = str_extract(sample_id, "^[0-9]+"),
    timepoint = if_else(str_detect(sample_id, "_"), str_extract(sample_id, "(?<=_)[0-9]+"), NA_character_)
  )

head(dengueIgG_serotype)
tail(dengueIgG_serotype)
```

### calculate baseline reactivity
```{r}
baseline <- dengueIgG_serotype |> 
  group_by(patient_id, serotype, probe_sequence) |> 
  filter(timepoint == min(timepoint)) |> 
  summarize(baseline_reactivity = mean(fluorescence_norm, na.rm = TRUE),
            .groups = 'drop')
```

### merge basline reactivity
```{r}
dengueIgG_serotype <- dengueIgG_serotype |> 
  left_join(baseline, by = c("patient_id", "serotype", "probe_sequence"))
```

### calculate change in reactivity
```{r}
dengueIgG_serotype <- dengueIgG_serotype |>
  mutate(delta_reactivity = fluorescence - baseline_reactivity)
```

### define patient-specific threshold
```{r}
threshold <- dengueIgG_serotype |> 
  group_by(patient_id, serotype) |> 
  summarize(
    threshold = 2 * sd(baseline_reactivity, na.rm = TRUE),
    .groups = 'drop'
  )

dengueIgG_serotype <- dengueIgG_serotype |>
  left_join(threshold, by = c("patient_id", "serotype"))
```

### mark significant changes
```{r}
dengueIgG_serotype <- dengueIgG_serotype |>
  mutate(significant_change = ifelse(abs(delta_reactivity) > threshold, 1, 0))
```

### calculate intensity and breadth of change
```{r}
varscore_components <- dengueIgG_serotype |> 
  group_by(patient_id, serotype, timepoint) |> 
  summarize(
    intensity_of_change = sum(delta_reactivity[significant_change == 1], na.rm = TRUE),
    breadth_of_change = mean(significant_change, na.rm = TRUE),
    .groups = 'drop'
  )
```

### normalize intensity and breadth across timepoints for each patient/serotype
```{r}
varscore_components <- varscore_components |> 
  group_by(patient_id, serotype) |> 
  mutate(
    norm_intensity = intensity_of_change / max(abs(intensity_of_change), na.rm = TRUE),
    norm_breadth = breadth_of_change / max(breadth_of_change, na.rm = TRUE)
  ) |> 
  ungroup()
```

### handle NAN for proper visualization
```{r}
varscore_components <- varscore_components %>%
  mutate(across(where(is.numeric), ~ifelse(is.nan(.), 0, .)))
```

### calculate time-adjusted VARscore
```{r}
varscore_components <- varscore_components |> 
  mutate(VARscore = 0.7 * norm_intensity + 0.3 * norm_breadth)
```



